\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-07 main.tex
small changes in read-through

\Abstract{
Spatio-temporal variables can be time invariant or not and this divides
them into those can be identified only with the spatial identifier and
those cross-identified by the spatial and temporal identifier. In this
paper, a new data structure, \code{cubble}, is proposed for manipulating
and visualising spatio-temporal data. The new data structure proposes
two forms for manipulating the spatial and temporal dimension separately
while providing a smooth switch between the two forms to link the two
dimensions. Advanced considerations are given to illustrate data with
hierarchical structure, data matching, and how \pkg{cubble} fits with
interactive graphics pipeline. Examples are given to analysing Australia
meteorology data, river level data, and climate reanalysis (ERA5).
}

@@

Spatio-temporal data concerns variables in the spatial coordinates
across time. Some variables are invariant to time and can be solely
identified by its spatial location while other time varying variables
need both the spatial and temporal identifier to identify. Manipulating
data in these two levels is usually done separately and the third tidy
data principal \citep{tidydata} recommends each type of observational
unit to form a table. Another feature for a spatio-temporal data
structure to have is that variables created from subsequent operations
can be automatically added to the existing table with the same level.
This is better than an output that index the variables with the
identifiers since otherwise analysts will need to manually join this
output with existing variables in the same level.

Currently, available spatio-temporal data structure in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses is based on \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts}, both of which has been replaced by
more recent implementations. \pkg{spatstat} \citep{spatstat} implements
a \texttt{ppp} class for point pattern data; and more recent,
\pkg{stars} \citep{stars} implements a spatio-temporal array with the
dplyr's data cube structure \pkg{cubelyr} \citep{cubelyr} as its
back-end. While these implementations either store spatial and temporal
variables all in a single table, hence duplicate the spatial variables
for each temporal unit; or split them into two separate tables that has
the problem of manually joining, mentioned in the previously. This
creates a gap in the software development. The requirement for such a
tool is important given the ubiquity of spatio-temporal vector data in
the wild: the Ireland wind data from \pkg{gstat} is an classic example
data that splits variables into its spatial (\texttt{wind.loc}) and
temporal (\texttt{wind}) dimension; Bureau of Meteorology (BoM) provides
climate observations widely applied in agriculture and ecology study.

This paper describes the implementation of a new spatio-temporal data
structure: \pkg{cubble}. \pkg{cubble} implements a relational data
structure that uses two forms to manage the switch between spatial and
temporal dimension. With this structure, users can manipulate the
spatial or temporal dimension separately, while leaves the linking of
two dimensions to \pkg{cubble}. The software is available from the
Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper will be divided as follows: Section 2 presents the
workflow of data manipulation in cubble. Section 3 explains how cubble
deals with some advanced considerations including data with hierarchical
structure, data matching, and how cubble fits with existing static and
interactive visualisation tools. Section 4 gives examples of the
features introduced in the previous two sections with climate and
hydrology data. An example on how cubble handle NetCDF data is also
provided. Section 5 concludes the paper.

@@

In a cubble, there are two forms: nested form and long form, and Figure
\ref{fig:illu-cubble} sketches the two forms along with the associated
attributes. The decision on which form to use is output-oriented,
meaning analysts need to first think about whether the output of a
particular operation is identified only by the spatial identifier, or a
combination of spatial and temporal identifier. The nested cubble is
suitable for working with operations that are only identified by the
site and this type of operation can be a pure manipulation of spatial
variables, or a summary of temporal variables by site (i.e.~the output
of counting the number of raining day is only identified by sites and
hence should be performed with the nested form). Underneath the nested
form, a cubble is built from a row-wise dataframe (\code{rowwise_df})
where each site forms a separate group. This structure simplifies the
calculation that involves temporal variables by avoiding the use of
\code{map} syntax when working with list-column.

For those operations whose output involves both a spatial and temporal
dimension, long form should be used. The long form is identified by both
the spatial and temporal identifier and adopts a grouped dataframe
(\texttt{grouped\_df}) to forms each site as a group. Spatial variables
are stored separately in a \pkg{tibble} as an special attribute of the
long cubble. This design avoids repeating the spatial variables at each
time stamp while not dropping information from spatial variables.

@@

\code{as_cubble()} creates a cubble by supplying the three key
components: \code{key} as the spatial identifier; \code{index} as the
temporal identifier; and a vector of \code{coords} in the order of
longitude first and then latitude. The naming of \code{key} and
\code{index} follows the convention in the \pkg{tsibble} package. The
cubble created by default is in the nested form. Below is an example of
creating a cubble:

@@

There are a few information in the cubble header: the name of the
\code{key} variable (\texttt{id}) and its number (5), bounding box
(\texttt{{[}115.97,\ -32.94,\ 133.55,\ -12.42{]}}), and also the name of
variable nested in the \code{ts} column with its type:
\code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

@@

The long cubble is suitable to manipulate the time dimension of the
data. The function \code{stretch()} switches the nested cubble into the
long cubble by first extracts all the spatial variables into a separate
tibble and store in the \code{spatial} attribute and then unnests the
\code{ts} column:

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_long <- cubble_nested %>% stretch(ts)
R> cubble_long
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

Notice here that the third line in the header now shows the name and
type of spatial variables:
\code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}.

@@

Spatial locations can have grouping structure that is either inherent to
the data (i.e.~countries are nested in continents), or formed by
clustering. Rather than analysing variables in the site level,
summarised variables in the cluster level can give a crisper picture of
local areas. In cubble, \code{switch_key()} can be used to create a new
level of grouping of spatial location by specifying a clustering
variable. Figure \ref{fig:illu-hier} illustrates the relationship of
cubbles at station and cluster level, in both the long and nested form.
By specifying
\code{cluster_nested <- station_nested \%>\% switch_key(key = cluster)},
the cubble re-defines the cubble key from \code{id} in
\code{station_nested} to \code{cluster} in \code{cluster_nested}. All
the spatial variables variant to \code{cluster} are now nested into a
\code{.val} column and cluster level variables can be computed in the
same fashion as station level variables in \code{station_nested}.
Following the same principal of \code{stretch}. the long form cluster
level cubble expands the time series variables with both indices:
\code{cluster} and \code{id}. All the cluster level variables are stored
separately for recovering the nested from from the long form.

@@

\code{match_sites()} from \pkg{cubble} matches data based on these two
criteria. It first matches the two data sources spatially through
computing the pairwise distance based on latitude and longitude. Pairs
that pass the spatial matching are then matched temporally through
computing the number of matched peaks within a fixed length window.
Figure \ref{fig:illu-matching} illustrate this temporal matching in more
details. Given two series \code{A} and \code{a}. 3 peaks have been
picked in each series. An interval, with default length of 5, is
constructed for each peak in series \code{A} and the peaks in series
\code{a} are tested against whether they fall into the any of the
intervals. In this illustration, there are 2 matches for these two
series. Several arguments are available in \code{match_sites()} to
fine-tune the matching:

@@

\begin{itemize}
\tightlist
\item
  \code{spatial_n_keep}. the number of spatial match for each site to
  keep
\item
  \code{spatial_dist_max}. the maximum distance allowed for a matched
  pair
\item
  \code{temporal_n_highest}. the number of peak used - 3 in the example
  above
\item
  \code{temporal_window}. the length of the interval - 5 in the example
  above
\item
  \code{temporal_min_match}. the minimum number of matched peak for a
  valid matched pair
\end{itemize}

This functionality can be seen as a matching
\citep{stuart2010matching, mcintosh2018using} exercise in the
spatio-temporal domain to pair up nearby observations. It can also be
considered as a form of data fusion
\citep{castanedo2013review, cocchi2019data}, where data from multiple
sources are combined together.

@@

Cubble fits in naturally with the interactive graphic pipeline discussed
in the literature
\citep{buja1988elements, buja1996interactive, sutherland2000orca, xie2014reactive, cheng2016enabling}.
Diagram \ref{fig:illu-interactive} illustrates how linking works from
the map to the time series with the two forms of a cubble. The map and
time series plot is associated with the nested or long cubble
respectively and when a user action is captured on the map, the site
will be activated in the nested cubble (left). The nested cubble will
communicate to the long cubble to activate all the observations with the
same \code{id} (middle). The long cubble will then highlight the
activated series in the time series plot (right).

@@

Glyph map \citep{Wickham2012-yr} plots the time series as single glyph
on the map and in \proglang{R}, \pkg{GGally} implements the glyph map
through the \code{glyphs()} function. The function construct a data
frame with calculated position (\code{gx}. \code{gy}. \code{gid}. of
each point on the time series using linear algebra (Equation 1 and 2 in
\citet{Wickham2012-yr}). The data can then be piped into \code{ggplot}
to create the glyph map as:

@@

Global Historical Climatology Network (GHCN) provides daily climate
measures from stations across the world and the dataset
\code{weatherdata::historical_tmax} extracts the maximum temperature for
236 Australian stations. All the stations start recording from year 1969
with the longest back to year 1859. The data \code{historical_tmax} is
already presented as a cubble, with \code{id} as the key, \code{date} as
the index, and \code{c(longitude, latitude)} as the coordinates. This
example compares the maximum temperature in two periods: 1971 - 1975 and
2016 - 2020 for stations in Victoria and New South Wales.

@@

In the previous example, there has already been some overlapping of the
glyphs for a few stations near (151E, 34S) and (152E, 33S) and this will
be a problem when mapping more stations in the national level.
Aggregation can be helpful to group series into clusters before
visualising the cluster with glyph map.

This example shows how to organise data at both level with
\code{switch_key()}. The dataset used is
\code{weatherdata::climate_full}, which records daily precipitation and
maximum/ minimum temperature for 639 stations in Australia from 2016 to
2020. A simple kmean algorithm based on the distance matrix is used here
to create 20 clusters. This creates \code{station_nested} as a station
level nested cubble with a cluster column indicating the group each
station belongs to. More complex algorithms can be used for other
problem, as long as there is a mapping from each station to a cluster.

@@

Bureau of Meteorology also collects
\href{http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full}{water
data} from river gauges, which includes electrical conductivity,
turbidity, water course discharge, water course level, and water
temperature. In particular, water level will interactive with
precipitation from the climate data since rainfall will raise the water
level in the river. Figure \ref{fig:matching-map} gives the location of
available weather station and water gauges in Victoria.

@@

NetCDF (Network Common Data Form) is a data format commonly used in
climatology community to deliver global mapping of atmosphere, ocean,
and land. It has two major components: dimension to define the
spatio-temporal grid (longitude, latitude, and time) and variable which
populates the defined grid. Attributes are usually associated with
dimension and variable in the NetCDF format data and a
\href{http://cfconventions.org/}{metadata convention for climate and
forecast} has been designed to standardise this data format. Current
packages on manipulating NetCDF data in R include a high-level R
interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls
C-interface: \pkg{RNetCDF} \citep{rnetcdf, michna2013rnetcdf}, and a
tidyverse implementation: \pkg{tidync} \citep{tidync}.

Cubble provides an \code{as_cubble()} method to coerce the \code{ncdf4}
class from the \pkg{ncdf4} package into a \code{cubble}. It maps each
combination of longitude and latitude into an \code{id} (the \code{key})
and nests time and variables in the nested form:

@@

Memory limit with NetCDF data in cubble depends on longitude grid point
x latitude grid point x time grid point x number of variable. Cubble can
handle slightly more than 300 x 300 (longitude x longitude) grid points
for 3 variables in one year and spatial grid can be reduced to trade for
longer time period and more variables. A 300 by 300 spatial grid can be
a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or
global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
Subsetting longitude and latitude grid is available through
\code{long_range} and \code{lat_range} if the NetCDF file has finer
resolution than needed.

@@

Figure \ref{fig:netcdf} gives an example of reproducing Figure 19 in
\citet{hersbach2020era5} using ERA5 data. ERA5 data
\citep{hersbach2020era5} is the latest reanalysis of global atmosphere,
land surface, and ocean waves from 1950 onwards and is available in the
NetCDF format. It can be directly downloaded from the
\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels?tab=overview}{Copernicus
Climate Data Store (CDS)} website or programmatically via R package
\pkg{ecmwfr} \citep{ecwmfr}. Variable Specific humidity and geopotential
are queried on the 10 hPa pressure level for four dates: 2002-09-22,
2002-09-26, 2002-09-30, and 2002-10-04 and once downloaded, the data can
be read into a cubble using the code above. Readers who are interested
in the details of can refer to \citet{simmons2020global} and
\citet{simmons2005ecmwf}.

@@

With spatio-temporal data, users may wish to make plots to learn the
spatial distribution of a variable or to find patterns such as trend or
seasonality in the time series. Combining this two types of plot with
interactivity let users to link between points on the map and their
corresponding time series and explore the spatial and temporal dimension
of the data simultaneously. Below is an example that describes the
process of building an interactive graphic with \pkg{cubble} and
\pkg{crosstalk}.

Starting with the original data, some pre-processing may be required to
summarise the data before the visualisation. This example explores the
variation of monthly temperature range across 639 weather stations in
Australia and with \texttt{weatherdata::climate\_full}, daily
temperature range is first calculated as the difference between
\code{tmax} and \code{tmin}. The three daily variables are then averaged
into month over 2016 - 2020 in the long form. Variance of the
temperature difference is then calculated for each station in the nested
form. Then, a \code{SharedData} object is constructed for each form of
the cubble and the same \code{group} argument ensures the cross-linking
of the two forms via the common \code{id} column. The spatial map and
time series plot are then made with each \code{SharedData} objects
separately. In this example, stations on the Australia map, made from
the nested form, are coloured by the calculated variance and a ribbon
band is constructed using the long form cubble to show the maximum and
minimum temperature of each station across month. With a different
dataset, users are free to calculate any per station measure in the
nested form or to make any time-wise summary of the data in the long
form to customise the spatial or temporal view. And the cross-linking is
always safeguarded by the shared \code{id} column embedded in the cubble
structure. Below is the pseudo code that outlines the process to
construct an interactive graphic described above:

@@

Figure \ref{fig:interactive-popup} shows the same content as Figure
\ref{fig:interactive-linking} but made with leaflet and popups.

@@

\caption[Same as FIgure 11 with thetemperature variation shown as a popup in the leaflet map]{Same as FIgure 11 with thetemperature variation shown as a popup in the leaflet map.}\label{fig:interactive-popup}
\end{figure}
\end{CodeChunk}

@@

This paper describes the \proglang{R} package \pkg{cubble} for
manipulating and visualising spatio-temporal data. A new data structure,
that builds from the \code{rowwise_df} and \code{grouped_df} class in
the tidyverse ecosystem, is proposed to connect the spatial and temporal
variables in the spatio-temporal data. This design frees the data
analysts from spending time on organising variables of different
observational units. The data structure is also flexible to the
techniques and packages analysts use to analyse the data, for example,
in the matching example in section 4.3, users are free to use algorithms
from another package to cluster stations. Further development and
maintenance of the package involves responding to changes in the
tidyverse packages that \pkg{cubble} imports, in particular,
\pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. Another area for further
development is to extend \pkg{cubble} to other spatial objects other
than points.

>>>>

\Abstract{
Spatio-temporal variables can be divided into time invariant ones that
can be identified by the spatial identifier, or the time varying ones
that cross-identified by the spatial and temporal identifiers. In this
paper, a new data structure, \code{cubble}, is proposed for manipulating
and visualising spatio-temporal data. The new data structure uses two
forms to organise time invariant/ varying variables so that the spatial
and temporal dimension of the data can be manipulated separately while
keep synchronised. Advanced considerations are given to illustrate data
with hierarchical structure, data matching, and how cubble fits with the
interactive graphics pipeline. Examples are given to analysing Australia
meteorology data, river level data, and climate reanalysis (ERA5).
}

@@

Spatio-temporal data concerns variables in the spatial coordinates
across time. Some variables are invariant to time and can be solely
identified by its spatial location while other time varying variables
need both the spatial and temporal identifiers to identify. Manipulating
data in these two levels is usually done separately and the third tidy
data principal \citep{tidydata} recommends each type of observational
unit to form a table. Another feature for a spatio-temporal data
structure to have is that variables created from subsequent operations
can be automatically added to the existing table with the same level.
This is better than an output that index the variables with the
identifiers since otherwise analysts will need to manually join this
output with existing variables in the same level.

Currently, available spatio-temporal data structures in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). \pkg{spacetime} is built on \pkg{sp} \citep{sp} and
\pkg{xts} \citep{xts}, both of which have been replaced by more recent
implementations. \pkg{spatstat} \citep{spatstat} implements a
\texttt{ppp} class for point pattern data; and more recent, \pkg{stars}
\citep{stars} implements a spatio-temporal for vector and raster
spatio-temporal data. While these implementations either store spatial
and temporal variables all in a single table, hence duplicate the
spatial variables at each temporal unit; or split them into two separate
tables that analysts need to manually join the data constantly when
doing the analysis. This creates a gap in the software development. The
requirement for such a tool is important given the ubiquity of
spatio-temporal vector data in the wild: climate observations from
weather stations, air quality variables from air monitoring stations, or
any variables that are measured at fixed location across time.

This paper describes the implementation of a new spatio-temporal data
structure: \pkg{cubble}. \pkg{Cubble} implements a data structure that
uses two forms to manage the switch between spatial and temporal
dimension. With this structure, users can manipulate the spatial or
temporal dimension separately, while leaves the linking of two
dimensions to \pkg{cubble}. The software is available from the
Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section 2 presents the
workflow of data manipulation in cubble. Section 3 explains how cubble
deals with some advanced considerations including data with hierarchical
structure, data matching, and how cubble fits with existing static and
interactive visualisation tools. Section 4 gives examples of the
features introduced in the previous two sections with climate and
hydrology data. An example on how cubble handle NetCDF data is also
provided. Section 5 concludes the paper.

@@

In a cubble, there are two forms: nested form and long form, and Figure
\ref{fig:illu-cubble} sketches the two forms with the associated
attributes. The decision on which form to use in data manipulation is
output-oriented, meaning analysts need to first think about whether the
output of an operation is identified only by the spatial identifier, or
a combination of spatial and temporal identifier. The nested cubble is
suitable for working with operations that are only identified by the
site and this type of operation can be a pure manipulation of time
invariant variables, or an operation that summarises time varying
variables into site. Underneath the nested form, a cubble is built from
a \code{rowwise_df} data frame where each site forms a separate group.
This structure simplifies the calculation that involves temporal
variables by avoiding the use of \code{purrr::map()} syntax when working
with list-column.

For those operations whose output involves both a spatial and temporal
identifier, long form should be used. The long form uses a
\code{`grouped_df`} data frame to forms all the time of a site as a
group. Time invariant variables are stored separately as an special
attribute of the long cubble. This design avoids repeating the spatial
variables at each time stamp while not dropping information from spatial
variables.

@@

A cubble can be created with \code{as_cubble()} by supplying the three
key components: \code{key} as the spatial identifier; \code{index} as
the temporal identifier; and a vector of \code{coords} in the order of
longitude and latitude. The naming of \code{key} and \code{index}
follows the convention in the \pkg{tsibble} package. Below is an example
of creating a cubble and by default, the cubble created is in the nested
form:

@@

There are a few information in the cubble header: the name of the
\code{key} variable (\texttt{id}) and its number (5), the bounding box:
\code{[115.97, -32.94, 133.55, -12.42]} , and also the name of variable
nested in the \code{ts} column with its type:
\code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

@@

The long cubble is suitable to manipulate the time dimension of the data
and \code{stretch()} switches the nested cubble into the long cubble. It
first extracts all the spatial variables into a separate tibble and then
unnests the \code{ts} column. The spatial variables are then attached as
an attributes of the long cubble and can be accessed through
\texttt{spatial()}

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_long <- cubble_nested %>% stretch(ts)
R> cubble_long
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\begin{CodeInput}
R> spatial(cubble_long)
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 5 x 6
  id            lat  long  elev name           wmo_id
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610
2 ASN00010311 -31.9  117. 179   york            94623
3 ASN00010614 -32.9  117. 338   narrogin        94627
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120
5 ASN00015131 -17.6  134. 220   elliott         94236
\end{CodeOutput}
\end{CodeChunk}

Notice that the third line in the header now shows the name and type of
spatial variables:
\code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}.

@@

Spatial locations can have grouping structure either inherent to the
data or formed by clustering. Rather than analysing variables in the
site level, summarised variables in the cluster level can give a crisper
picture of local areas. In cubble, \code{switch_key()} can be used to
create a new level of grouping of spatial locations by specifying a
clustering variable. Figure \ref{fig:illu-hier} illustrates the
relationship of cubbles at station and cluster level, in both the long
and nested form. By specifying
\code{cluster_nested <- station_nested \%>\% switch_key(key = cluster)},
the cubble re-defines the cubble key from \code{id} in
\code{station_nested} to \code{cluster} in \code{cluster_nested}. All
the spatial variables variant to \code{cluster} are now nested into a
\code{.val} column and cluster level variables can be computed in the
same fashion as station level variables in \code{station_nested}.

@@

\code{match_sites()} matches data based on these two criteria: it first
matches the two data sources spatially through computing the pairwise
distance on latitude and longitude. Pairs that pass the spatial matching
are then matched temporally through computing the number of matched
peaks within a fixed length window. Figure \ref{fig:illu-matching}
illustrates this temporal matching in more details. Given two series
\code{A} and \code{a}, 3 peaks have been picked in each series. An
interval, with default length of 5, is constructed for each peak in
series \code{A} and the peaks in series \code{a} are tested against
whether they fall into the any of the intervals. In this illustration,
there are 2 matches for these two series. Several arguments are
available in \code{match_sites()} to fine-tune the matching:

@@

\begin{itemize}
\tightlist
\item
  \code{spatial_n_keep}: the number of spatial match for each site to
  keep
\item
  \code{spatial_dist_max}: the maximum distance allowed for a matched
  pair
\item
  \code{temporal_n_highest}: the number of peak used - 3 in the example
  above
\item
  \code{temporal_window}: the length of the interval - 5 in the example
  above
\item
  \code{temporal_min_match}: the minimum number of matched peak for a
  valid matched pair
\end{itemize}

This functionality can be seen as a matching exercise
\citep{stuart2010matching, mcintosh2018using} in the spatio-temporal
domain to pair up nearby observations. It can also be considered as a
form of data fusion \citep{castanedo2013review, cocchi2019data}, where
data from multiple sources are combined together.

@@

Cubble fits in naturally with the interactive graphic pipeline discussed
in the literature
\citep{buja1988elements, buja1996interactive, sutherland2000orca, xie2014reactive, cheng2016enabling}.
Diagram \ref{fig:illu-interactive} illustrates how linking works from
the map to the time series in cubble. The map and time series plot is
associated with the nested or long cubble, respectively, and when a user
action is captured on the map, the site will be activated in the nested
cubble (left). The nested cubble will communicate to the long cubble to
activate all the observations with the same \code{id} (middle). The long
cubble will then highlight the activated series in the time series plot
(right).

@@

Glyph map \citep{Wickham2012-yr} plots the time series as single glyph
on the map and in \proglang{R}, \pkg{GGally} implements the glyph map
through the \code{glyphs()} function. The function constructs a data
frame with calculated position (\code{gx}. \code{gy}. \code{gid}) of
each point on the time series using linear algebra (Equation 1 and 2 in
\citet{Wickham2012-yr}). The data can then be piped into \code{ggplot}
to create the glyph map as:

@@

Global Historical Climatology Network (GHCN) provides daily climate
measures from stations across the world. The dataset
\code{weatherdata::historical_tmax} extracts the maximum temperature for
236 Australian stations from GHCN with starting from year 1969.
\code{weatherdata::historical_tmax} is already in a cubble, with
\code{id} as the key, \code{date} as the index, and
\code{c(longitude, latitude)} as the coordinates. This example compares
the maximum temperature in two periods: 1971 - 1975 and 2016 - 2020 for
stations in Victoria and New South Wales.

@@

In the previous example, there has already been some overlapping of the
glyphs for a few stations near (151E, 34S) and (152E, 33S) and this will
be a problem when mapping more stations in the national level.
Aggregation can be helpful to group series into clusters before
visualising the cluster with glyph map. This example shows how to
organise data at both level with \code{switch_key()}.

\code{weatherdata::climate_full}, also extracted from GHCN, records
daily precipitation and maximum/ minimum temperature for 639 stations in
Australia from 2016 to 2020. A simple kmean algorithm based on the
distance matrix is used to create 20 clusters. This creates
\code{station_nested} as a station level nested cubble with a cluster
column indicating the group each station belongs to. More complex
algorithms can be used for other problem, as long as there is a mapping
from each station to a cluster.

@@

Bureau of Meteorology collects
\href{http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full}{water
data} from river gauges and this includes variables: electrical
conductivity, turbidity, water course discharge, water course level, and
water temperature. In particular, water level will interactive with
precipitation from the climate data since rainfall will raise the water
level in the river. Figure \ref{fig:matching-map} gives the location of
available weather station and water gauges in Victoria.

@@

NetCDF (Network Common Data Form) is a data format commonly used in the
climatology community to deliver global mapping of atmosphere, ocean,
and land. It has two major components: \emph{dimension} for defining the
spatio-temporal grid (longitude, latitude, and time) and \emph{variable}
that populates the defined grid. Attributes are usually associated with
dimension and variable in the NetCDF format data and a
\href{http://cfconventions.org/}{metadata convention for climate and
forecast} has been designed to standardise the format of the attributes.
Current packages for manipulating NetCDF data in R include a high-level
R interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls
C-interface: \pkg{RNetCDF} \citep{rnetcdf, michna2013rnetcdf}, and a
tidyverse implementation: \pkg{tidync} \citep{tidync}.

Cubble provides an \code{as_cubble()} method to coerce the \code{ncdf4}
class from the \pkg{ncdf4} package into a \code{cubble}. It maps each
combination of longitude and latitude into an \code{id} as the
\code{key}:

@@

Memory limit with NetCDF data in cubble depends on longitude grid point
x latitude grid point x time grid point x number of variable. Cubble can
handle slightly more than 300 x 300 (longitude x longitude) grid points
for 3 variables in one year and spatial grid can be reduced to trade for
longer time period and more variables. A 300 by 300 spatial grid can be
a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or a
global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
Subsetting longitude and latitude grid is available through
\code{long_range} and \code{lat_range} if the NetCDF file has finer
resolution than needed.

@@

Figure \ref{fig:netcdf} gives an example of reproducing Figure 19 in
\citet{hersbach2020era5} using ERA5 data. ERA5 data
\citep{hersbach2020era5} is the latest reanalysis of global atmosphere,
land surface, and ocean waves from 1950 onwards and is available in the
NetCDF format from European Centre for Medium-Range Weather Forecasts
(ECMWF). The data can be directly downloaded from
\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels?tab=overview}{Copernicus
Climate Data Store (CDS)} website or programmatically via an R package
\pkg{ecmwfr} \citep{ecwmfr}. Variable \emph{specific humidity} and
\emph{geopotential} are queried on the 10 hPa pressure level on four
dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04. Once
downloaded, the data can be read into a cubble as:

\begin{CodeChunk}
\begin{CodeInput}
R> raw <- ncdf4::nc_open(here::here("data/era5-pressure.nc"))
R> dt <- as_cubble(raw, vars = c("q", "z"))
\end{CodeInput}
\end{CodeChunk}

Readers who are interested in the analysis of the figure can refer to
\citet{simmons2020global} and \citet{simmons2005ecmwf} for more details.

@@

With spatio-temporal data, users may wish to make plots to learn the
spatial distribution of a variable, or to find patterns, such as trend
or seasonality, in the time series. Combining this two types of plot
with interactivity let users to link between points on the map and the
corresponding time series to explore the spatial and temporal dimension
of the data simultaneously. Below is an example that describes the
process of building an interactive graphic with \pkg{cubble} and
\pkg{crosstalk} The example explores the variation of monthly
temperature range with \code{weatherdata::climate_full} data.

The temperature range is calculated as the difference between
\code{tmax} and \code{tmin} and its monthly average over 2016 - 2020 is
taken before calculating the variance. A \code{SharedData} object is
constructed for each form of the cubble and the same \code{group}
argument ensures the cross-linking of the two forms via the common
\code{id} column. The spatial map and time series plot are then made
with each \code{SharedData} objects separately. In this example,
stations on the Australia map, made from the nested form, are coloured
by the calculated variance and a ribbon band is constructed using the
long form cubble to show the maximum and minimum temperature of each
station across month. With a different dataset, users are free to
calculate any per station measure in the nested form or to make any
time-wise summary of the data in the long form to customise the spatial
or temporal view. The cross-linking between the two plots is always
safeguarded by the shared \code{id} column embedded in the cubble
structure. Below is the pseudo code that outlines the process to
construct an interactive graphic described above:

@@

Figure \ref{fig:interactive-popup} shows Figure
\ref{fig:interactive-linking} made made with leaflet and popups.

@@

\caption[Same as Figure 11 with thetemperature variation shown as a popup in the leaflet map]{Same as Figure 11 with thetemperature variation shown as a popup in the leaflet map.}\label{fig:interactive-popup}
\end{figure}
\end{CodeChunk}

@@

This paper describes an \proglang{R} package \pkg{cubble} for
manipulating and visualising spatio-temporal data. A new data structure,
\code{cubble} that builds from the \code{rowwise_df} and
\code{grouped_df} class in the tidyverse ecosystem, is proposed to
connect the time invariant and varying variables in the spatio-temporal
data. This design frees the data analysts from spending time on
organising variables of different observational units. The data
structure is also flexible to the techniques and packages analysts use
to analyse the data, for example, in the matching example in section
4.3, users are free to use algorithms from another package to cluster
stations. Further development and maintenance of the package involves
responding to changes in the tidyverse packages that \pkg{cubble}
imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}.
Another area for further development is to extend \pkg{cubble} to other
spatial objects other than points.
